---
title: "Herongate regression modelling"
author: "David Wachsmuth"
date: '2022-05-26'
output:
  word_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# NB! A censusmapper.ca API key needs to be specified, e.g. with:
# cancensus::set_api_key(KEY_HERE)

```

## Attach packages

```{r packages}

library(tidyverse)
library(sf)
library(cancensus)
library(tongfen)
library(patchwork)

```

## Construct vectors of 2006/2016 census variables

For convenience, we pre-specify the vectors of census variables for the 2006 and 2016 censuses, using common variable names. (These were retrieved using `cancensus::list_census_vectors`.)

```{r vectors}

# 2006 vector
vec_2006 <- c(
  population = "v_CA06_1",
  total_students = "v_CA06_1248",
  degree = "v_CA06_1254",
  income = "v_CA06_2000",
  total_children = "v_CA06_78",
  u17_1 = "v_CA06_79",
  u17_2 = "v_CA06_80",
  u17_3 = "v_CA06_81",
  total_marital = "v_CA06_41",
  married = "v_CA06_43",
  total_language = "v_CA06_243",
  english = "v_CA06_244",
  french = "v_CA06_245",
  bilingual = "v_CA06_246",
  total_lf = "v_CA06_575",
  lf = "v_CA06_576",
  employed = "v_CA06_577",
  unemployed = "v_CA06_578",
  poverty = "v_CA06_1981",
  rooms = "v_CA06_99",
  total_built = "v_CA06_109",
  built1 = "v_CA06_110",
  built2 = "v_CA06_111",
  built3 = "v_CA06_112",
  built4 = "v_CA06_113",
  built5 = "v_CA06_114",
  built6 = "v_CA06_115",
  built7 = "v_CA06_116",
  built8 = "v_CA06_117",
  built9 = "v_CA06_118",
  total_houses = "v_CA06_119",
  detached = "v_CA06_120",
  semidetached = "v_CA06_121",
  row = "v_CA06_122",
  duplex = "v_CA06_123", # This was previously called "apartment"
  big_apartment = "v_CA06_124",
  small_apartment = "v_CA06_125",
  rent = "v_CA06_2050",
  value = "v_CA06_2054",
  occupied_dwellings = "v_CA06_101",
  owned = "v_CA06_102",
  rented = "v_CA06_103",
  band_housing = "v_CA06_104",
  total_repairs = "v_CA06_105",
  minor1 = "v_CA06_106",
  minor2 = "v_CA06_107",
  major = "v_CA06_108",
  total_race = "v_CA06_1302",
  indigenous = "v_CA06_565",
  vismin = "v_CA06_1303",
  chinese = "v_CA06_1304",
  southasian = "v_CA06_1305",
  black = "v_CA06_1306",
  filipino = "v_CA06_1307",
  latin = "v_CA06_1308",
  southeast = "v_CA06_1309",
  arab = "v_CA06_1310",
  westasian = "v_CA06_1311",
  korean = "v_CA06_1312",
  japanese = "v_CA06_1313",
  notvismin = "v_CA06_1316"
  )

# 2016 vector
vec_2016 <- c(
  population = "v_CA16_401",
  total_students = "v_CA16_5096",
  degree = "v_CA16_5105", # Previously had v_CA16_5117 (non-university certificate)
  income = "v_CA16_2397", # Previously had v_CA16_2398 (after-tax hh income)
  total_children = "v_CA16_2510", # Not comparable to 2006 children variable
  under17 = "v_CA16_2513", # Not comparable to 2006 children variable
  total_marital = "v_CA16_451",
  married = "v_CA16_457", # Previously had v_CA16_454 (married & common law)
  total_language = "v_CA16_512",
  english = "v_CA16_515",
  french = "v_CA16_518",
  bilingual = "v_CA16_521",
  total_lf = "v_CA16_5597",
  lf = "v_CA16_5600",
  employed = "v_CA16_5603",
  unemployed = "v_CA16_5606",
  poverty = "v_CA16_2570",
  rooms = "v_CA16_4855",
  total_built = "v_CA16_4862",
  prior1960 = "v_CA16_4863",
  prior1980 = "v_CA16_4864",
  prior1990 = "v_CA16_4865",
  prior2000 = "v_CA16_4866",
  prior2006 = "v_CA16_4867",
  prior2011 = "v_CA16_4868",
  prior2016 = "v_CA16_4869",
  total_houses = "v_CA16_408",
  detached = "v_CA16_409",
  semidetached = "v_CA16_412",
  row = "v_CA16_413",
  duplex = "v_CA16_414",
  big_apartment = "v_CA16_410", 
  small_apartment = "v_CA16_415",
  rent = "v_CA16_4901",
  value = "v_CA16_4896",
  occupied_dwellings = "v_CA16_4836",
  owned = "v_CA16_4837",
  rented = "v_CA16_4838",
  band_housing = "v_CA16_4839",
  total_repairs = "v_CA16_4870",
  minor = "v_CA16_4871",
  major = "v_CA16_4872",
  total_race = "v_CA16_3954",
  indigenous = "v_CA16_3855",
  vismin = "v_CA16_3957",
  chinese = "v_CA16_3963",
  southasian = "v_CA16_3960",
  black = "v_CA16_3966",
  filipino = "v_CA16_3969",
  latin = "v_CA16_3972",
  southeast = "v_CA16_3978",
  arab = "v_CA16_3975",
  westasian = "v_CA16_3981",
  korean = "v_CA16_3984",
  japanese = "v_CA16_3987",
  notvismin = "v_CA16_3996"
  )

# Give them names which will make it easy to split years and variables later
names(vec_2006) <- paste(names(vec_2006), "2006", sep = "___")
names(vec_2016) <- paste(names(vec_2016), "2016", sep = "___")

```

## Build dataset

We start by downloading the variables from `vec_2006` and `vec_2016` using the `tongfen` package, which constructs a common geometry for multiple census periods. We specify the CMA codes for Montreal, Ottawa, Toronto, Calgary, Edmonton and Vancouver—the six CMAs with more than 1 million population in 2016—`level = "CT"` to retrieve the data at the census tract level, and `base_geo = "CA16"` to build the common geographies from 2016 census tracts.

```{r tongfen}

CT_raw <- get_tongfen_ca_census(
  regions = list(CMA = c("35535", "24462", "59933", "48825", "505", "48835")), 
  meta = meta_for_ca_census_vectors(c(vec_2006, vec_2016)),
  level = "CT",
  base_geo = "CA16",
  quiet = TRUE)

```

Next we separate out the tongfen geometries for future use.

``` {r geometries}

CT_geom_raw <- 
  CT_raw |> 
  select(TongfenUID, geometry)

```

### Clean dataset

Now we start to clean up the dataset, first by adding a density variable, then pivoting the table by year and then sorting out the variables that were handled differently in the different census years. We also update `CT_geom` and remove the geometry from the main `CT` dataset.

``` {r cleanup_1}

CT <- 
  CT_raw |> 
  mutate(
    density___2006 = units::drop_units(population___2006 / st_area(geometry)),
    density___2016 = units::drop_units(population___2016 / st_area(geometry))) |>
  pivot_longer(-c(TongfenID:geometry),
               names_to = c("variable", "year"),
               names_sep = "___") |> 
  st_drop_geometry() |> 
  pivot_wider(names_from = variable, values_from = value) |> 
  relocate(under17, .after = total_children) |> 
  mutate(under17 = coalesce(under17, u17_1 + u17_2 + u17_3)) |>
  select(-u17_1, -u17_2, -u17_3) |> 
  mutate(prior1960 = coalesce(prior1960, built1 + built2),
         prior1980 = coalesce(prior1980, built3 + built4),
         prior1990 = coalesce(prior1990, built5 + built6),
         prior2000 = coalesce(prior2000, built7 + built8),
         prior2006 = coalesce(prior2006, built9)) |> 
  relocate(prior1960:prior2016, .after = total_built) |> 
  select(-c(built1:built9)) |> 
  mutate(minor = coalesce(minor, minor1 + minor2)) |> 
  select(-minor1, -minor2) |> 
  relocate(minor, .after = major) |> 
  left_join(CT_geom_raw, by = "TongfenUID") |> 
  rename(census_tract = TongfenID) |> 
  select(-TongfenUID) |> 
  mutate(city = substr(census_tract, 1, 3), .after = census_tract) |> 
  mutate(city = case_when(
    city == "462" ~ "Montreal",
    city == "505" ~ "Ottawa",
    city == "535" ~ "Toronto",
    city == "825" ~ "Calgary",
    city == "835" ~ "Edmonton",
    city == "933" ~ "Vancouver")) |> 
  st_as_sf()

CT_geom <- 
  CT |> 
  select(census_tract, geometry)

CT <- 
  CT |> 
  st_drop_geometry()

```

We now continue cleanup by removing CTs with low population or incomplete coverage, and add various computed variables.

``` {r cleanup_2}

CT <- 
  CT |> 
  # Drop CTs with low population
  filter(!is.na(population), population >= 100) |> 
  # Drop CTs only present in one year
  group_by(census_tract) |> 
  filter(n() == 2) |> 
  ungroup() |> 
  # Create percentage variables
  mutate(p_college = degree / total_students,
         p_children = under17 / total_children, # Previously was under17 / population
         p_detached = detached / total_houses,
         p_racialized = (vismin + indigenous) / total_race,
         p_black = black / total_race,
         p_white = (notvismin - indigenous) / total_race)

CT <- 
  CT |> 
  # Create change in rent, value, income and p_racialized for 2016
  group_by(census_tract) |> 
  mutate(rent_change = rent - rent[year == "2006"], 
         rent_change_pct = rent_change / rent[year == "2006"],
         value_change = value - value[year == "2006"], 
         value_change_pct = value_change / value[year == "2006"],
         income_change = income - income[year == "2006"],
         income_change_pct = income_change / income[year == "2006"],
         racialized_change = p_racialized - p_racialized[year == "2006"]) |> 
  ungroup()

```

We pivot the table again to structure it properly for the regression models.

``` {r pivot}

CT_final <-
  CT |> 
  pivot_wider(names_from = year, values_from = population:racialized_change) |> 
  select(-value_change_2006, -value_change_pct_2006, -rent_change_2006,
         -rent_change_pct_2006, -income_change_2006, -income_change_pct_2006, 
         -racialized_change_2006) |> 
  rename(value_change = value_change_2016,
         value_change_pct = value_change_pct_2016,
         rent_change = rent_change_2016,
         rent_change_pct = rent_change_pct_2016,
         income_change = income_change_2016,
         income_change_pct = income_change_pct_2016,
         racialized_change = racialized_change_2016)

```

We add the new Black "location quotient" variable

``` {r black_var}

CT_final <- 
  CT_final |> 
  mutate(black_ratio_2006 = p_black_2006 / 
           (sum(p_black_2006 * total_race_2006, na.rm = TRUE) / 
              sum(total_race_2006, na.rm = TRUE)),
         black_ratio_2016 = p_black_2016 / 
           (sum(p_black_2016 * total_race_2016, na.rm = TRUE) / 
              sum(total_race_2016, na.rm = TRUE)))

CT_final <- 
  CT_final |> 
  mutate(black_ratio_change = black_ratio_2016 - black_ratio_2006)

```

### Add distance variable

We now add distance to city hall (or a similar relevant city-centre location) to act as a spatial control in the models.

``` {r distance_var}

ch_montreal <- 
  st_point(c(-73.55426429176865, 45.50887112946624)) |> 
  st_sfc(crs = 4326)
  
ch_ottawa <- 
  st_point(c(-75.70055444964218, 45.42343521835774)) |> 
  st_sfc(crs = 4326)

ch_toronto <- 
  st_point(c(-79.38364973603952, 43.65276898078397)) |> 
  st_sfc(crs = 4326)

ch_calgary <- 
  st_point(c(-114.0563303787967, 51.04631058985368)) |> 
  st_sfc(crs = 4326)

ch_edmonton <- 
  st_point(c(-113.49036821589088, 53.544862921879975)) |> 
  st_sfc(crs = 4326)

ch_vancouver <- 
  st_point(c(-123.1182923162272, 49.28260752585029)) |> 
  st_sfc(crs = 4326)

ch <- c(ch_montreal, ch_ottawa, ch_toronto, ch_calgary, ch_edmonton, 
        ch_vancouver) |> 
  st_as_sf() |> 
  mutate(city = c("Montreal", "Ottawa", "Toronto", "Calgary", "Edmonton",
                  "Vancouver"), .before = x) |> 
  rename(geometry = x)

ch_dist <- 
  CT_geom |> 
  distinct(census_tract, .keep_all = TRUE) |> 
  rowwise() |> 
  mutate(ph_dist = min(as.numeric(st_distance(geometry, ch)))) |> 
  ungroup() |> 
  st_drop_geometry()

CT_final <- 
  CT_final |> 
  left_join(ch_dist, by = "census_tract")

```

### Scale variables for regression models

Finally, we scale all variables for the regression models

``` {r scale}

scale_fun <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)

CT_final <- 
  CT_final |> 
  mutate(across(where(is.numeric), ~if_else(is.infinite(.x), NA_real_, .x))) |> 
  mutate(across(where(is.numeric), ~if_else(is.nan(.x), NA_real_, .x))) |> 
  mutate(across(where(is.numeric), scale_fun)) |> 
  mutate(city = as.factor(city))

```


## Regression models

All models are linear regressions with CMA acting as a fixed effect. (I.e. each CMA will have a separate intercept to account for inter-regional variation in the outcome variable which isn't otherwise captured by the model's predictors.)

### 2016 owner-occupied housing value

``` {r value_2016}

model_2016_value_fe <- 
  lm(value_2016 ~ black_ratio_2016 + income_2016 + detached_2016 +
       rooms_2016 + p_children_2016 + p_college_2016 + 
       employed_2016 + density_2016 + poverty_2016 +
       ph_dist + city - 1, 
       data = CT_final)

summary(model_2016_value_fe)

```


### 2016 owner-occupied housing value (per-city)

``` {r value_2016_per_city}

model_2016_value_fe_list <- 
  map(unique(CT_final$city), ~{
  lm(value_2016 ~ black_ratio_2016 + income_2016 + detached_2016 +
       rooms_2016 + p_children_2016 + p_college_2016 + 
       employed_2016 + density_2016 + poverty_2016 +
       ph_dist, 
     data = filter(CT_final, city == .x))
}) |> set_names(unique(CT_final$city))

map(model_2016_value_fe_list, summary)

```

### 2006-2016 value change

``` {r value_change}

model_value_change_fe <- 
  lm(value_change_pct ~ value_2006 + black_ratio_2016 + black_ratio_change + 
       income_2016 + detached_2016 + rooms_2016 + p_children_2016 + 
       p_college_2016 + employed_2016 + density_2016 + poverty_2016 +
       ph_dist + city - 1, 
     data = CT_final)

summary(model_value_change_fe)

```


### 2006-2016 value change (per-city)

``` {r value_change_per_city}

model_value_change_fe_list <- 
  map(unique(CT_final$city), ~{
  lm(value_change_pct ~ value_2006 + black_ratio_2016 + black_ratio_change + 
       income_2016 + detached_2016 + rooms_2016 + p_children_2016 + 
       p_college_2016 + employed_2016 + density_2016 + poverty_2016 +
       ph_dist, 
     data = filter(CT_final, city == .x))
    }) |> set_names(unique(CT_final$city))

map(model_value_change_fe_list, summary)

```


### 2016 rent

``` {r rent_2016}

model_2016_rent_fe <- 
  lm(rent_2016 ~ black_ratio_2016 + income_2016 + detached_2016 +
         rooms_2016 + p_children_2016 + p_college_2016 + ph_dist + city - 1, 
       data = CT_final)

summary(model_2016_rent_fe)

```

### 2006-2016 change in rent

``` {r rent_change}

model_rent_change_fe <- 
  lm(rent_change ~ value_2006 + black_ratio_2016 + black_ratio_change + 
       income_2016 + rooms_2016 + detached_2016 + p_college_2016 + 
       p_children_2016 + ph_dist + city - 1, data = CT_final)

summary(model_rent_change_fe)

```

## Basic visualizations

These are extremely barebones visualizations of the key predictor and outcome variables in each of the models. We can work these up to be properly presentable when necessary.

### 2016 Black population

``` {r black_2016_map}

# Assemble list of plots
black_2016_maps <- 
  map(c("Calgary", "Edmonton", "Ottawa", "Montreal", "Toronto", "Vancouver"), ~{
    CT_final |> 
      filter(city == .x) |> 
    select(census_tract, black_ratio_2016) |> 
    left_join(distinct(CT_geom, census_tract, .keep_all = TRUE), 
              by = "census_tract") |>
    st_as_sf(crs = 4326) |> 
    ggplot(aes(fill = black_ratio_2016)) +
    geom_sf(colour = "transparent") +
    scale_fill_viridis_b(limits = c(-2, 2), oob = scales::squish,
                         n.breaks = 7) +
      ggtitle(.x) +
    theme_void() +
    theme(legend.position = "right")
})

# Lay out plots with patchwork
do.call(wrap_plots, c(black_2016_maps, list(guides = "collect", ncol = 2)))
        
```


### 2006-2016 change in Black population

``` {r black_change_map}

# Assemble list of plots
black_change_maps <- 
  map(c("Calgary", "Edmonton", "Ottawa", "Montreal", "Toronto", "Vancouver"), ~{
    CT_final |> 
      filter(city == .x) |> 
    select(census_tract, black_ratio_change) |> 
    left_join(distinct(CT_geom, census_tract, .keep_all = TRUE), 
              by = "census_tract") |>
    st_as_sf(crs = 4326) |> 
    ggplot(aes(fill = black_ratio_change)) +
    geom_sf(colour = "transparent") +
    scale_fill_viridis_b(limits = c(-2, 2), oob = scales::squish,
                         n.breaks = 7) +
      ggtitle(.x) +
    theme_void() +
    theme(legend.position = "right")
})

# Lay out plots with patchwork
do.call(wrap_plots, c(black_change_maps, list(guides = "collect", ncol = 2)))
        
```

### 2016 value

``` {r value_2016_map}

# Assemble list of plots
value_maps <- 
  map(c("Calgary", "Edmonton", "Ottawa", "Montreal", "Toronto", "Vancouver"), ~{
    CT_final |> 
      filter(city == .x) |> 
    select(census_tract, value_2016) |> 
    left_join(distinct(CT_geom, census_tract, .keep_all = TRUE), 
              by = "census_tract") |>
    st_as_sf(crs = 4326) |> 
    ggplot(aes(fill = value_2016)) +
    geom_sf(colour = "transparent") +
    scale_fill_viridis_b(limits = c(-2, 2), oob = scales::squish,
                         n.breaks = 7) +
      ggtitle(.x) +
    theme_void() +
    theme(legend.position = "right")
})

# Lay out plots with patchwork
do.call(wrap_plots, c(value_maps, list(guides = "collect", ncol = 2)))
        
```

### 2006-2016 change in value

``` {r value_change_map}

# Assemble list of plots
value_change_maps <- 
  map(c("Calgary", "Edmonton", "Ottawa", "Montreal", "Toronto", "Vancouver"), ~{
    CT_final |> 
      filter(city == .x) |> 
    select(census_tract, value_change) |> 
    left_join(distinct(CT_geom, census_tract, .keep_all = TRUE), 
              by = "census_tract") |>
    st_as_sf(crs = 4326) |> 
    ggplot(aes(fill = value_change)) +
    geom_sf(colour = "transparent") +
    scale_fill_viridis_b(limits = c(-2, 2), oob = scales::squish,
                         n.breaks = 7) +
      ggtitle(.x) +
    theme_void() +
    theme(legend.position = "right")
})

# Lay out plots with patchwork
do.call(wrap_plots, c(value_change_maps, list(guides = "collect", ncol = 2)))
        
```


### 2016 rent

``` {r rent_2016_map}

# Assemble list of plots
rent_maps <- 
  map(c("Calgary", "Edmonton", "Ottawa", "Montreal", "Toronto", "Vancouver"), ~{
    CT_final |> 
      filter(city == .x) |> 
    select(census_tract, rent_2016) |> 
    left_join(distinct(CT_geom, census_tract, .keep_all = TRUE), 
              by = "census_tract") |>
    st_as_sf(crs = 4326) |> 
    ggplot(aes(fill = rent_2016)) +
    geom_sf(colour = "transparent") +
    scale_fill_viridis_b(limits = c(-2, 2), oob = scales::squish,
                         n.breaks = 7) +
      ggtitle(.x) +
    theme_void() +
    theme(legend.position = "right")
})

# Lay out plots with patchwork
do.call(wrap_plots, c(rent_maps, list(guides = "collect", ncol = 2)))
        
```

### 2006-2016 change in rent

``` {r rent_change_map}

# Assemble list of plots
rent_change_maps <- 
  map(c("Calgary", "Edmonton", "Ottawa", "Montreal", "Toronto", "Vancouver"), ~{
    CT_final |> 
      filter(city == .x) |> 
    select(census_tract, rent_change) |> 
    left_join(distinct(CT_geom, census_tract, .keep_all = TRUE), 
              by = "census_tract") |>
    st_as_sf(crs = 4326) |> 
    ggplot(aes(fill = rent_change)) +
    geom_sf(colour = "transparent") +
    scale_fill_viridis_b(limits = c(-2, 2), oob = scales::squish,
                         n.breaks = 7) +
      ggtitle(.x) +
    theme_void() +
    theme(legend.position = "right")
})

# Lay out plots with patchwork
do.call(wrap_plots, c(rent_change_maps, list(guides = "collect", ncol = 2)))
        
```
